Follow-up Samples

### 2017-02-23 submission/manuscript_outline_20161024.Rmd

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e. adenoma or carcinoma) removal.  Specifically, we wanted to identify what within the community was different before and after removal of said lesion.

**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` while a model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  Follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.  The initial sample model had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

@@

Over the last few years a few studies have shown how this bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as specific Operational Taxonomic Units (OTUs) relative abundance.  We utilized Random Forest to build two models.  The first model was built to classify lesion and normal samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.

@@

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal controls.  We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].  The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].  Although there was a detectable change and this change was towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

***Study Design and Patient Sampling:***
The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to normalize within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after lesion removal and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

@@

***Lesion Model Creation:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on a data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.

@@

***Initial Sample Model Creation:***
We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal samples from each other.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial to lesion removal sample.

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was appropriate a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.  Finally, a list of common OTUs were found for the two two different models used.

@@

**Figure 1: General Differences between the Adenoma and Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). C) NMDS of the initial and follow up samples for the Adenoma group.  D) NMDS of the initial and follow up samples for the Carcinoma group. For C) and D) the teal represents initial samples and the pink represents follow up samples.

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with previous research suggesting that a link may exist between the fecal bacterial microbiome and CRC.  The overall objective of our study was to test the hypothesis that changes in the bacterial microbiome occur after lesion (i.e. adenoma or carcinoma) removal.  Specifically, we wanted to identify components within the community that were different before and after removal of said lesion.

**Results:** The bacterial microbiome changed more in response to lesion removal in carcinoma cases compared to adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference for either the adenoma or carcinoma group in the relative abundance of any OTU between pre and post lesion removal (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A model built to classify lesion had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had a decrease in the positive probability of lesion (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05) suggesting a movement towards a more normal bacterial community.  A second model built to classify initial samples had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease in positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

@@

Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma or carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion (adenoma or carcinoma).  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models.  The first model was built to classify lesion and normal (non-lesion) samples while the second was built to classify initial and follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial and follow up.

@@

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma and carcinoma versus normal controls.  We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].  The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial (before lesion was removed) samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].

@@

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after lesion removal.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma and carcinoma).  Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].  Although there was a detectable change towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 3C & 3D].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

@@

***Study Design and Patient Sampling:***
The sampling and design were similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.  Samples used to build the models for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to normalize back to a pre-colonscopy community within this time period [@obrien_impact_2013].  Our follow up data set had a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after lesion removal and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

@@

***Lesion Model Creation:***
The Random Forest [@breiman_random_2001] algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) with the main training and testing of the model completed on an independent data set of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals.  This model was then applied to our follow up data set of `r length(rownames(good_metaf))` individuals.  It should be noted that all individuals with an adenoma or carcinoma were grouped together to form the lesion group and the model was not created to find differences between normal, adenoma, and carcinoma but rather differences between both adenoma and carcinoma versus normal.

@@

***Initial Sample Model Creation:***
We also investigated whether a model could be created that could identify pre (initial) and post (follow up) lesion removal samples from each other.  The main difference was that only the `r length(rownames(good_metaf))`-person cohort was used at all stages of model building and classification.  Other than this difference the creation of this model and optimization of the mtry hyper-parameter was completed using the same procedure as was used for the lesion model.  Instead of classifying samples as positive or negative of lesion this model classified samples as positive or negative for being an initial sample prior to lesion removal.

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between initial and follow up samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

***Analysis Overview:***
We first wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.  This was done by testing initial and follow up samples for differences in alpha and beta diversity, testing differences in FIT between initial and follow ups, testing all OTUs, and investigating the relative abundance of specific previously associated CRC bacteria (*Fusobacterium nucleatum*, *Parvimonas micra*, *Peptostreptococcus assacharolytica*, and *Porphyromonas stomatis*) based on adenoma and carcinoma.  From here the lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  We then used the initial sample model to assess whether it could classify samples better then the lesion model and whether it could reduce the positive probability of an initial sample in the follow up samples.  Finally, a list of common OTUs were found for the two different models used.

@@

**Figure 1: General Differences between the Adenoma and Carcinoma Group.**  A) A significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).  B) A significant difference was found between the adenoma and carcinoma group for change in FIT measurement (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`). Advanced adenomas are denoted as Screen Relevant Neoplasia (SRN).  C) NMDS of the initial and follow up samples for the adenoma group.  D) NMDS of the initial and follow up samples for the carcinoma group.
